---
title: "Indiscriminate Mass Shootings in the U.S."
author: "Jamison R. Crawford, MPA"
date: "August 7, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '4'
---
<style>
body {
text-align: justify}
</style>

# Overview & Data Sources

The following visualizes indiscriminate mass shootings in the United States since 1982. 

These data derive from an [open source database](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/) assembled by U.S. magazine *Mother Jones*, which defines "indiscriminate rampages" as having four or more victims and which "exclude[s] shootings stemming from more conventionally motivated crimes such as armed robbery or gang violence" ([Follman et al. 2019](https://www.motherjones.com/politics/2012/07/mass-shootings-map/)).

Instead, these massacres address the influences of these growing phenomena, "from the firearms used and mental health factors to the growing copycat problem" (ibid.).

<br>

```{r echo=FALSE, message=FALSE, eval=FALSE, warning=FALSE}
    # Date: 2019-08-07
    # R Version: 3.6.0
    # RStudio Version: 1.2.1335
    # Operating System: Windows 10
    # Source: Follman, Mark et al. (2019). 
        # "US Mass Shootings, 1982-2019: Data From Mother Jones’ Investigation". 
        # Mother Jones. Retrieved on 07 August 2019 from
        # https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/
```

```{r echo=FALSE, message=FALSE, eval=TRUE, warning=FALSE}

# Install & Load Packages

if(!require(readr)){install.packages("readr")}
if(!require(dplyr)){install.packages("dplyr")}
if(!require(stringr)){install.packages("stringr")}
if(!require(leaflet)){install.packages("leaflet")}
if(!require(lubridate)){install.packages("lubridate")}
if(!require(kableExtra)){install.packages("kableExtra")}

library(readr)
library(dplyr)
library(stringr)
library(leaflet)
library(lubridate)
library(kableExtra)

# Read In Data (Source: Mother Jones, )

url <- paste0("https://docs.google.com/spreadsheets/d/e",
              "/2PACX-1vQBEbQoWMn_P81DuwmlQC0_jr2sJDzkk",
              "C0mvF6WLcM53ZYXi8RMfUlunvP1B5W0jRrJvH-wc",
              "-WGjDB1/pub?gid=0&single=true&output=csv")

mass <- read_csv(url)

rm(url)

```

```{r echo=FALSE, message=FALSE, eval=TRUE, warning=FALSE}

# Dimension Reduction, Class Coersion, & Cleaning

mass <- mass %>%
  select(case:total_victims, sources, weapon_type:weapon_details, 
         race:gender, latitude:longitude) %>%
  mutate(date = mdy(date),
         case = str_to_title(case),
         fatalities = as.integer(fatalities),
         injured = as.integer(injured),
         total_victims = as.integer(total_victims))

# Converting State Abbreviations

mass$city <- str_split(mass$location, pattern = ", ", simplify = TRUE)[, 1]
mass$state <- str_split(mass$location, pattern = ", ", simplify = TRUE)[, 2]

mass <- left_join(mass, tibble(state = state.abb,
                               state_full = state.name), 
                  by = "state")

index <- which(!is.na(mass$state_full))

mass[index, "state"] <- mass[index, "state_full"]

mass <- mass %>%
  mutate(location = paste(city, state, sep = ", ")) %>%
  select(-city:-state_full)

# Convert URLs to Hyperlinks

mass <- mass %>%
  mutate(sources = str_replace_all(sources, pattern = "; .*$", ""),
         link = paste0("<a href='", sources, "'>", case, "</a>"))

# Title & Sentence Case; Full Gender

mass$weapon_type <- str_to_title(mass$weapon_type)
mass$weapon_details <- str_to_title(mass$weapon_details)

mass$race <- str_to_title(mass$race)
mass$race <- str_replace_all(mass$race, pattern = "^-$", "Unavailable")

mass$gender <- str_replace_all(mass$gender, pattern = "^M$", "Male")
mass$gender <- str_replace_all(mass$gender, pattern = "^F$", "Female")

names(mass) <- str_replace_all(names(mass), "_", " ")
names(mass) <- str_to_title(names(mass))  

```

# Visualizing Massacres

While similar to the original [interactive map](https://www.motherjones.com/politics/2012/07/mass-shootings-map/) by *Mother Jones*, the popup data provided in the below map (`Map I`) includes additional data, including article links, total injured victims and fatalities, city and state, weapons and weapon details, e.g. modifications, and both race and gender of the shooter(s).

**How to Use this Map****:** Each circle in the below map represents one mass shooting, while the size of the circle indicates total fatalities per incident. Use your mouse wheel or the `+` or `-` icons to zoom in and out. Click and drag the map to move the window in different directions. Click a circle to view the details of the shooting.

<br>

<center>

**Map I: Indiscriminate Mass Shootings in the U.S.**

```{r echo=FALSE, message=FALSE, eval=TRUE, warning=FALSE, fig.width=8.5, fig.height=6}
mass %>%
  leaflet() %>%
  addTiles() %>%
  addCircleMarkers(lng = mass$Longitude,
                   lat = mass$Latitude,
                   weight = 1.25,
                   radius = mass$Fatalities,
                   color = "tomato", 
                   opacity = 1, 
                   fillOpacity = 0.5,
                   popup = paste(mass$Link, 
                                 "<br>", mass$Location,
                                 "<br>", mass$Date,
                                 "<br><br>Perpetrator Race:", mass$Race,
                                 "<br>Perpetrator Gender:", mass$Gender,
                                 "<br><br>Injuries: ", mass$Injured,
                                 "<br>Fatalities: ", mass$Fatalities,
                                 "<br>Total Victims: ", mass$`Total Victims`,
                                 "<br><br>Weapon(s): ", mass$`Weapon Type`,
                                 "<br><br>Weapon Details: ", mass$`Weapon Details`, 
                                 "<br><br>", mass$Summary))
```

*Larger circles indicate more fatalities. Zoom in to more easily view each mass shooting.* 

</center>

<br>

# Appendices

The following table (`Table I`) contains key variables related to the mass shootings visualized above.

<center>

<br>

**Table I: Mass Shooting Names, Places, Dates, and Victim Count**

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}

appn <- mass %>%
  select(Case:Date, `Total Victims`) %>%
  arrange(desc(`Total Victims`))

kable(appn) %>%
  kable_styling(bootstrap_options = c("striped", 
                                      "hover", 
                                      "responsive"),
                font_size = 14,
                full_width = TRUE)

```

</center>

<br>

# References

**Follman, Mark, Gavin Aronsen, Deanna Pan (2019).** "A Guide to Mass Shootings in America". *Mother Jones*. Retrieved on 07 August 2019 from [https://www.motherjones.com/politics/2012/07/mass-shootings-map/](https://www.motherjones.com/politics/2012/07/mass-shootings-map/).

**Follman, Mark, Gavin Aronsen, Deanna Pan (2019).** "US Mass Shootings, 1982 - 2019: Data From Mother Jones' Investigation". *Mother Jones*. Retrieved on 07 August 2019 from [https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/).

<br>

# Disclaimer

The use of *Mother Jones* as a data source, as well as the subject matter, do not necessarily reflect on the sociopolitical views or ideologies of the analyst and author of the present work.